Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-19 paper/paper.qmd
Tweak perf caption; notes

A few ways to define CDI severity (@fig-flowchart)

@@

Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).

@@

@@

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the Infectious
Diseases Society of America (IDSA) based on lab values,
the occurence of an adverse outcome due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to
CDI with chart review (Attrib).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

::: {#fig-performance fig-cap="**Performance of ML models.**
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection set.
<!-- TODO add plots of AUROC and AUBPRC curves -->
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection set.
OTU 120 (*Pseudomonas*) is not shown for the full data set with IDSA severity 
on the full dataset because it was removed during pre-processing due to having
near-zero variance.
**B)** Log~10~ transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape).
"}
![](figures/feature-importance.png)
:::

>>>>

prediction models based on EHR for whether infection occurs in the first place already in use.
so how about predicting severity of infections to guide treatment?
OTUs vs EHRs to predict severity.

A few ways to define CDI severity (@fig-flowchart).

@@

Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.

@@

NNT x NNS = NNB

NNT for Fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is Vancomycin
because it's cheaper than Fidaxomicin, even though IDSA recommends Fidaxomicin.

@@

{{< pagebreak >}}

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the Infectious
Diseases Society of America (IDSA) based on lab values,
the occurrence of an adverse outcome due to any cause (All-cause),
and the occurrence of disease-related complications confirmed as attributable to
CDI with chart review (Attrib).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

::: {#fig-performance fig-cap="**Performance of ML models.**
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. 
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus an identical performance is expected.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
<!-- TODO add plots of AUROC and AUBPRC curves -->
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
OTU 120 (*Pseudomonas*) is not shown for IDSA severity on the full dataset nor
on the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape).
"}
![](figures/feature-importance.png)
:::
